---
title: "Salmon Project"
author: "Team Data Driven"
date: "Deisharrah Allen-Benso, Jonathan Acevedo, Reuben Lopez, Rachel Tekchandani"
output: pdf_document
---


# Introduction.
In this project we investigate contamination levels in farm raised salmon. The obtained data set consists of 153 instances of farm raised salmon. It includes salmons' farms origin and 11 different contaminants found in salmon. We can split our data into three regions: North America, South America and Europe. In this research we would like to focus on which region of farm salmon has a higher mean Mirex contaminant. Additionally we are interested to see if some contaminants are independent from other contaminants in their respective levels. To address these questions, we will use statistical methods like ANOVA(analysis of variance), permutation based ANOVA and permutation test of independence of two variables. 
There are papers that also focused salmon contamination levels. For example, 'Global Assessment of Organic Contaminants in Farmed Salmon. (2003)'. However, above mentioned paper and others were focusing on mean contamination locations in particular locations rather than by regions. Additionally, none of them focused on any form of correlation between contaminants. 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```



```{r readFile, include=FALSE}
file.choose()  # choose the directry
data = read.csv('salmonData.csv', header =TRUE,)
```



# Results and discussion.
Our data analysis starts with data exploration. Results of data analysis can be seen in the Appendix section of this paper. We can observe that Mirex column contains has three NaNs values. Hence, we omit them from further analysis. This makes our data set to contain 150 rows going forward. 
We can also observe that there are 8 unique locations in our dataset.



```{r}
data = na.omit(data)
unique( data$Location )
```

$~$

We will define the three regions using the following guidelines:


North America - Eastern Canada,  Western Canada,Washington, Maine.


South America - Chile.


Europe - Norway, Scotland and Faroe Islands


We will encode these three regions as 1,2, 3 respectively.

$~$

```{r}
data$Region = 1
data$Region[data$Location %in% c('Norway', 'Scotland','Faroe Islands')] = 3
data$Region[data$Location %in% c('Chile')] = 2

```



Let's see number of examples available by Region.

```{r}

regionDF <- data.frame("Region" = c('North America','South America', 'Europe'), "Instances" = c(57, 30, 63))
print(regionDF)
#knitr::kable(regionDF, format = "pdf")
```


Let's see assess correlation coefficients between Mirex contaminant and all other contaminants.

```{r}
contaminants = colnames(data)[4:length(data)-1]
cor_values = array(length(contaminants))
i = 1
for( i in 1:length(contaminants))
{
  #print(c_name)
  c_name = contaminants[i]
  #print(cor(data['Mirex'], data[c_name]))
  cor_values[i] = cor(data['Mirex'], data[c_name])

}
corrDataFrame <- data.frame("Contaminant" = contaminants, "Correlation" = cor_values)
print(corrDataFrame)
#knitr::kable(corrDataFrame, format = "pdf")

```


We can observe that all contaminants but HCH_gamma are highly correlated to Mirex. HCH_gamma contaminant has negative correlation of -0.028. We will investigate whether we have a small negative correlation or a correlation of 0 cannot be ruled out. Lets check whether we can reject the hypothesis that correlation between HCH_gamma and Mirex is statistically 0. Due to sample sizes we will employ the permutation test of independence of two variables. 




```{r, echo=FALSE, out.width="50%", out.height="50%", fig.cap="Bootstrap correlation distribution.", fig.align='center'}
N <- 10^5
n <- nrow(data) 
result <- numeric(N)
observed <- cor(data['Mirex'], data['HCH_gamma'])[1,1]
for (i in 1:N)
{
   index <- sample(n , replace = FALSE)
   HCH.permuted <- data$HCH_gamma[index]
   result[i] <-     cor(HCH.permuted, data$Mirex)
}

 
hist(result)
abline(v = observed, col = "blue")  
print('The p-value is')
print( (sum(result >= observed) + 1)/(N+1) )

```

Figure 1 contains bootstrap distribution of correlation values.The observed p-value is ~ 0.6. Since p-value is greater that 5% alpha, we fail to reject the null hypothesis that contaminants Mirex and HCH_gamma are independent.  



```{r, out.width="50%", out.height="50%", fig.cap="Box-plots of Mirex level per region", fig.align='center'}
boxplot(Mirex~Region,data=data, main="Mirex level",
   xlab="Region", ylab="Mirex level")
```


Lets check average Mirex levels for each region.  

```{r}
mean_val = array(3)
for( r in c(1,2,3))
{
mean_val[r] = mean( data[data$Region == r,'Mirex'])
}
regionDF <- data.frame("Region" = c('North America','South America', 'Europe'), "Mean" = mean_val)
print(regionDF)

```

Although average values seem to differ, we see from box-plots in Figure 2 that overall distributions are similar across 3 regions. Hence, to assess whether we can reject the null hypothesis that mean values between regions are different we employ the ANOVA test.   Let's define $\mu_1 , \mu_2 , \mu_3$ as means of Mirex contaminant values for three regions. Therefore, ANOVA test hypotheses look like this:  


$H_0:\mu_1 = \mu_2 = \mu_3$  
$H_A:$ means are not equal  

$~$

```{r}
Regions <- as.factor(data$Region)
g = lm(data$Mirex ~ Regions)
anova(g)

```


```{r}
TukeyHSD(aov(data$Mirex ~ Regions))

```

We can observe that at 5% significance level there is a difference between Regions 2 and 1 , 3 and 1, 3 and 2.
Our overall p-value is 2.2e-16. Hence, we have sufficient evidence to reject the null hypothesis that means of Mirex between 3 regions are the same.

Based on quantile plot of residuals of ANOVA table, we can see that residuals are not normally distributed. Hence, standard ANOVA assumption of residuals being normally distributed is violated. Therefore, it is better to proceed with permutation based version.


```{r, echo=FALSE, out.width="50%",out.height="50%",fig.cap="Bootstrap correlation distribution.", fig.align='center'}
Regions <- as.factor(data$Region)
g = lm(data$Mirex ~ Regions)
anova(g)
observed <- anova(g)$F[1]

n <- length(data$Mirex)
N <- 10^4 - 1
results <- numeric(N)
for (i in 1:N)
{
 index <- sample(n)
 Mirex.perm <- data$Mirex[index]
 results[i] <- anova(lm(Mirex.perm ~ Regions, data =data))$F[1]
}
hist(results)
abline(v = observed, col = "blue")  
print('Our p-value is')
print( (sum(results >= observed) + 1) / (N + 1) )

```

The permutation based p-value is $e$^-4. Hence, we reject the null hypothesis that average Mirex contaminant values are the same across 3 regions.


# Conclusion.
In our findings we have seen that Mirex contamination levels in Europe are higher than in other two regions. We have also observed that Mirex and HCH contaminants do not seem to have any linear association between them.






# References.

1. Chihara, L., & Hesterberg, T. (2019). Mathematical statistics with resampling and R. Hoboken, NJ: Wiley.
2. Global Assessment of Organic Contaminants in Farmed Salmon. (2003). Ronald A. Hites, Jeffery A. Foran, David O. Carpenter,
   M. Coreen Hamilton, Barbara A. Knuth, Steven J. Schwager
3. B. Charron, An IntraFish.com Industry Report on Salmon Product Development–The Fish of the Future? (1999) 
   
   
\newpage
# Appendix.
Summary of Salmon data.

```{r}
if (!require("Hmisc"))
  { install.packages("Hmisc") 
library('Hmisc')}
describe(data)

```

```{r}
Regions <- as.factor(data$Region)
g = lm(data$Mirex ~ Regions)
anova(g)
plot(g)[1]

```











